---
title: "Untitled"
author: "Gerard Castell i Joan Valls"
date: "Friday, March 20, 2015"
output:
  pdf_document:
    fig_caption: yes
---

## Multivariate model

```{r, echo=FALSE, results='hide'}
round.pval <- function(pval=stop("p-value not given"), show.sig.stars=FALSE, alpha=0.05){
  
  arrodonir <- function(pval, show.sig.stars, alpha){
		
		if(is.na(pval)) return(NA)

		ndec <- nchar(as.character(signif(alpha,1)))-2
		pval2 <- if(pval >= alpha) round(pval, ndec) else signif(pval, 1)

		i <- 1
		while(pval2==signif(alpha, i)){
			pval2 <- round(pval, ndec+i)
			if(pval2==pval) break
			i <- i+1
		}
		pval2 <- format(pval2, scientific=F)
		if(pval<0.00001){
			pval2 <- "<0.00001"
		}
		
		if(show.sig.stars){
			cpoints <- c(0, alpha/50, alpha/5, alpha, ceil(alpha*10^(ndec-1))/10^(ndec)+alpha, 1)
			c(0, 0.001, 0.01, 0.05, 0.1, 1)
			pval2 <- paste(pval2, symnum(pval, corr = FALSE, na = FALSE, cutpoints = cpoints, symbols = c("***", "**", "*", ".", " ")))
		}

		return(pval2)
	}

	ret <- sapply(pval, function(x) arrodonir(x, show.sig.stars=show.sig.stars, alpha=alpha))
	if(is.matrix(pval)) ret <- matrix(ret, ncol=ncol(pval))
	return(ret)

}

cows <- read.table("data/cattle_mes dades.txt", header = TRUE, 
                   sep = "\t", dec = ",", na.strings = "")

names(cows) <- c("id", "dose", "pcv", "time", "nbirth")

cows$id <- as.factor(cows$id)
cows$dose <- factor(cows$dose, levels=c("L", "M", "H"))

# dep variable:
# pcv: the higher the better.

# indep variables:
# dose: we want to know if the dose is associated with lower PCV
# nbirth: we want to know if this covariate influences the effect of the dose.
# time: we dont particularly want to see differences in time, but maybe 
# 3rd observations of the doses are higher, ...

# we do not want:
# id: we dont want the effect of the dose to be explained by the cow.

# Each observation of each cow will be considered independent.
taula <- matrix(ncol=6, nrow=19)
model <- lm(pcv ~ time + dose, data=cows)
summary(model)
model0 <- model
colnames(taula) <- c("Formula", "Variables", colnames(summary(model)$coef))
as.character(model$terms)->formu
taula[1,1] <- paste(formu[c(2,1,3)], collapse="")
taula[1:4,-(1:2)] <-  summary(model)$coef
taula[1:4,2] <- rownames(summary(model)$coef)

model <- lm(pcv ~ time*dose, data=cows)
summary(model)
model1 <- model
as.character(model$terms)->formu
taula[6,1] <- paste(formu[c(2,1,3)], collapse="")
taula[6:11,-(1:2)] <-  summary(model)$coef
taula[6:11,2] <- rownames(summary(model)$coef)
# dose H has a huge effect, in a possitive way. Also, in general higher
# times give better PCV values (which is logical). nbirth has a significative
# negative effect on the PCV.

model <- lm(pcv ~ time*dose + nbirth, data=cows)
summary(model)
model2 <- model
as.character(model$terms)->formu
taula[13,1] <- paste(formu[c(2,1,3)], collapse="")
taula[13:19,-(1:2)] <-  summary(model)$coef
taula[13:19,2] <- rownames(summary(model)$coef)
# time and dose almost interact for high doses. This interaction is synergistic.
# higher times and dose H give even higher pcv, but this effect is higher when 
# combined together.

model <- lm(pcv ~ dose + time*nbirth, data=cows)
summary(model)
as.character(model$terms)->formu
model <- lm(pcv ~ time + dose*nbirth, data=cows)
summary(model)
# no other interactions are significative.
taula[,6] <- round.pval(as.numeric(taula[,6]))
taula[,3:5] <- round(as.numeric(taula[,3:5]), 2) 
# print(xtable(taula, digits=c(0,0,0,2,2,2,2)), include.rownames=F)
# so... model selection, for the more pasimonious model says...
add1(model0, scope=~.^2+nbirth, test="F")
add1(update(model0,~.^2), scope=~.+nbirth, test="F")
add1(model2, scope=~.^2, test="F")

## final model:
model2 <- lm(pcv ~ time*dose + nbirth, data=cows, y=T, model=T)
summary(model2)
anova(model2)
```

Table \ref{taulamodels} summarizes the results of the estimation of the parameters of all models considered with the forward stepwise method for model selection. The last one corresponds to the most parsimonious model that better explains the data, whose especification is the following:

$$
PCV= 15.82 + 2.43X_{time} -0.32X_{doseM} -1.31X_{doseH} -0.34X_{nbirth} + 0.48X_{doseM}X_{time} + 2.35X_{doseH}X_{time}
$$

Although the coefficients for the additive effects for both doses are negative, when adding the interaction term it ensures a possitive effect. The difference in PCV for the same cows treated with high dose with respect to low dose will be:

$$
PCV(X_{doseH}=1)-PCV(X_{doseL}=1) = -1.31+2.35X_{time} 
$$

Notice that this is a positive quantity for every time, so we can consider that the high dose has a bigger effect than low dose (both coefficients significantly different than 0). For the medium dose, although being also positive, the coefficients are not significant, so could be 0. This calculations are always with respect to the same cow, when being treated with different doses.

$$
PCV(X_{doseM}=1)-PCV(X_{doseL}=1) = -0.32+0.48X_{time} 
$$

Therefore, the effect of the dose then changes with the time in a synergistic way. 
Also, time has a possitive effect in the PCV, as it goes by PCV grows (p$<$0.00001). However the number of births seems to have a significant negative effect in the PCV (p=0.01).



\begin{table}[ht]
\centering
\begin{tabular}{llllll}

\hline
Formula & Variables & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
pcv\~{}time + dose & (Intercept) & 12.32 & 0.83 & 14.78 & $<$0.00001 \\ 
   & time & 3.55 & 0.37 & 9.63 & $<$0.00001 \\ 
   & doseM & 0.56 & 0.75 & 0.75 & 0.45 \\ 
   & doseH & 3.2 & 0.72 & 4.47 & 0.00003 \\ 
   &  &  &  &  &  \\ 
  pcv\~{}time * dose & (Intercept) & 14.51 & 1.29 & 11.28 & $<$0.00001 \\ 
   & time & 2.28 & 0.68 & 3.34 & 0.001 \\ 
   & doseM & -0.47 & 1.81 & -0.26 & 0.79 \\ 
   & doseH & -1.45 & 1.73 & -0.84 & 0.41 \\ 
   & time:doseM & 0.65 & 0.93 & 0.7 & 0.48 \\ 
   & time:doseH & 2.5 & 0.87 & 2.87 & 0.005 \\ 
   &  &  &  &  &  \\ 
  pcv\~{}time * dose + nbirth & (Intercept) & 15.85 & 1.35 & 11.74 & $<$0.00001 \\ 
   & time & 2.43 & 0.66 & 3.67 & 0.0005 \\ 
   & doseM & -0.32 & 1.75 & -0.18 & 0.86 \\ 
   & doseH & -1.31 & 1.67 & -0.78 & 0.44 \\ 
   & nbirth & -0.34 & 0.13 & -2.51 & 0.01 \\ 
   & time:doseM & 0.48 & 0.9 & 0.54 & 0.59 \\ 
   & time:doseH & 2.35 & 0.84 & 2.8 & 0.007 \\ 
   \hline
\end{tabular}
\caption{All multivariant models considered for fitting the data that were significantly different, using F-tests to assess this difference.}
\label{taulamodels}

\end{table}

The analyses of the residuals of the model can be seen in Figure \ref{resplots}. Clearly, they present some issues: residuals are not independent from PCV, are not centered for high dose, no homoscedasticity, etc. So even thought this is not the optimal model to fit the data (and this is also shown in the residuals plots), the model is pointing in a pretty obvious way that the high dose is the only one that an effect significantly different from low dose. One of the problems with the data is that there are no controls among the cows so, the effect of the doses cannot be compared with PCV without any treatment.


```{r, echo=FALSE, fig.cap='Analyses of the residuals of the model.\\label{resplots}'}
## diagnostic:
par(mfrow=c(2,2), mar=c(4,4,2,2), cex=0.6)
# high pcv has high positive errors. Doesnt seem to be random.
plot(model2$residuals, model2$y, xlab="residuals", ylab="PCV")
# problems with doseH
plot(model2$model$dose, model2$residuals, ylab="residuals", xlab="Dose")
# NO homoscedasticity
plot(model2$model$time, model2$residuals, ylab="residuals", xlab="Time")
# not centered
plot(model2$model$nbirth, model2$residuals, ylab="residuals", xlab="Number of births (nbirth)")

## BAD MODEL: we already know that, since obervations are not independent.
```

