---
title: "Study of the efficacy of Berenil applied to trypanosomosis's infected cattle"
author: "Gerard Castellà, Mathieu Marauri and Lluis Ramon"
date: \today
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
header-includes:
- \usepackage{float}
---


```{r initial_chunk, echo = FALSE, warning = FALSE, message = FALSE}
library("knitr")
library("ggplot2")
library("xtable")
library("dplyr")
library("nlme")
library("gridExtra")
opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE)
```

```{r}
cows <- read.table("data/cattle_mes dades.txt", header = TRUE, 
                   sep = "\t", dec = ",", na.strings = "")

names(cows) <- c("id", "dose", "pcv", "time", "nbirth")
cows$dose <- factor(cows$dose, levels = c("L", "M", "H"))
cows$doseExt <- factor(cows$dose, labels = c("Low", "Medium", "High"))
```


# Introduction

Cattle are an important economic resource and also a major health factor in many countries. Therefore preventing diseases in a herd is vital. 

In this study we focus our interest on the parasite of trypanosomosis which can lead to the death of a cow. This disease, transmitted by the tsetse flies, causes an infection characterized by fever, loss of appetite and anemia, which can lead to death depending on different factors.

One medicine, the Berenil, is used to cure the infected cattle. The aim of this research is to determine the efficiency of different doses of Berenil. Finding the most efficient dose, if such a dose exists, is critical when it comes to save both cattle and money. Here lies all the interest of our study. 

To assess the severity of the disease, the Packed-Cell Volume (PCV) is shown to be a good indicator. The PCV is the percentage of red blood cells in blood. The lower the PCV is, the less healthy the cow is. 

# Objective

The aim of this study is to assess the efficacy of different doses of Berenil in cattle infected with the trypanosomosis parasite.

# Dataset

A cohort of 10 different cows infected by trypanosomosis parasite was selected for the study. Each  Berenil dose (low, medium and high) was administrated three times (time 1, 2 or 3) for each animal. PCV was reported each time as well as the number of calves it had before being infected.

The variables reported for this study are presented below.

* Id: Each cow has its own id. From 1 to 10. _(id)_
* PCV: From 14.30 to 33.30. _(pcv)_
* Dose: H High, M Medium L low. _(dose)_
* Time: From 1 to 3. _(time)_
* Number of birth: From 2 to 8. _(nbirth)_

Since the gathering process of the data was unknown, several assumptions were needed.

* Each time PCV is obtained before the treatment. Therefore the effect of the third dose could not be evaluated.
* Dose is assigned randomly in time to the cow.
* For a given cow, the previous treatments (high, medium or low) do not affect the following ones.
* Missing values are completely at random.
* Time intervals are the same and fixed.

# Statistical methods

The methods used in the statistical analyses are detailed in this section. First a multivariate model was fitted, then a two stage analysis and finally a random effect model. An exploratory data analysis was performed before those regression methods.

## Multivariate model

Despite the fact of not being the optimal choice, a multivariate linear model was fitted to the data. It seems reasonable to think that observations are not independent, since they are coming from the same cow. However, this model could also led to some interesting results.

The response variable was PCV since the severity the disease was evaluated by this parameter. As we wanted to estimate the effect of dose in diseased cows in time, these two variables were included in the model. 

From this initial model, a forward step-wise method was carried for the model selection, including additional covariates or interactions between them. The models were compared using F-test. This procedure was followed until no other covariates or interactions below 3rd order could be added, because they would not produce a significantly better model. Leading this to the most parsimonious model with all significant variables.

A validation of the model was performed to check that the residuals were normal, homoscedastic and independent between them and to other relevant covariates. 

## Two-stage analysis

A Two-stage analysis was performed to take into account variability between subjects.

In the first stage model, 3 linear regression models were done for each subject separately, one for each dose (High, Medium, Low).  The response variable  was PVC and time the covariate. 

In the second stage model, a linear regression model was fitted for the subject-specific regression coefficients using dose as a covariate. It was also studied the cow effect or the number of births effect.

## Random-effects model
\label{Statistical_methods:RE}

Since the Two Stage model didn't take into account the within subjects variability, a Random-effects model was fitted. 

It distinguishes the mean structure of the covariates as a fixed effect and also some variability between and within subjects as a random effects. It allows more flexibility to model.

In order to select the best model the following methodology was used.

1. Selection of a Preliminary Mean Structure
2. Selection of a Preliminary Random-Effects Structure
3. Selection of a Residual Covariance Structure

To compare different models the same estimation method was needed. In case of nested fixed effects a Maximum Likelihood method was used. Whenever there were random effects Restricted Maximum Likelihood was used. To chose the best model a Likelihood Ratio test was performed. If it was significant we kept the new variable or random effect. If not we drop it to obtain a more parsimonious model.

A validation of the model was performed to check that the residuals were normal, homoscedastic and independent between them and to other relevant covariates. 


# Summary statistics

In this part the aim is to have some insights about the data. By performing a basic statistical analysis some relations between the explaining variables and the response one would emerged. A first idea on how we can explain the response variable and also some answers when it comes to discovering the most efficient dose.

```{r qplot, fig.cap = "PCV evolution for each cow on every dose. Cows are colored in the same fashion for each dose. \\label{fig:qplot}", fig.pos = "H", fig.height = 3, fig.width = 7}
qplot(factor(time), pcv, data = cows, group = id, geom = "line", facets = .~ doseExt, 
      colour = factor(id)) + scale_color_discrete(guide = 'none') + xlab("time") + ylab("PCV value")
```


PCV values seem similar for each dose at time 1, which makes sense with several assumptions previously stated. Especially the fact that the PCV is taken before treatment. The unfinished evolutions in Figure \ref{fig:qplot} represent missing values in the data, more details about missing data are reported in the section \ref{sec:Missing values}. Notice that several cows with low PCV values in Low dose also have low PCV values in the other doses. 

Cows treated with high doses tend to have a better evolution in their PCV values. One could also think that time has an influence on the PCV in the way that the effect of the second dose seems to be less important than the first one. 


The mean influence of the covariate *time* is expected to be positive. Indeed one expects to higher PCV values after treatment. Also the High dose is expected to be more efficient than the Medium and Low one. This can be verified in Figure \ref{fig:boxplot_time_dose}.

```{r boxplot, fig.cap = "Boxplot of PCV values for each time and dose. \\label{fig:boxplot_time_dose}", fig.pos = "H", fig.height = 4, fig.width = 6}
cows$timeDose <- paste(cows$time, cows$dose, sep = ".")
cows$timeDose <- factor(cows$timeDose, levels = c("1.L", "2.L", "3.L", "1.M", "2.M", "3.M", "1.H", "2.H", "3.H"))
ggplot(aes(timeDose, pcv, fill = doseExt), data = cows) + geom_boxplot() +
  xlab("time") + ylab("PCV value") + scale_x_discrete(labels = c(1:3)) +
   theme(legend.title=element_blank())
cows$timeDose <- NULL
```

In Figure \ref{fig:boxplot_time_dose} it can be appreciated that there is not a big difference between the low and the medium doses. However the High dose is much better than the others. For instance the mean PCV value at time 2 for the High dose is bigger than mean PCV value at time 3 for the Medium or Low dose.

## Number of births

None of the cows gave birth after time 1 of the first infection. So the number of births remains constant through all doses and times for each cow. 

Some boxplots were performed but it didn't show any clear result, although a tendency may be seen at time 1. A simple regression model was also fit at time 1 with a significant (p = 0.007) slope  of -0.3444. It may explain the fact that a cow is supposed to be weaker if it had a lot of calves. Latter on the analysis the nbirth effect will be combined with other variables.

```{r, eval = FALSE}
lcowsnbirth <- lm(pcv ~ nbirth, data = cows[cows$time == 1,])
summary(lcowsnbirth)
```

## Missing data analysis
\label{sec:Missing values}

There are 14 missing PCV values out of 90 observations in the dataset. Those missing values are equally distributed among doses and time as it is shown in Table 1.

```{r, results = 'asis'}
cows.com <- na.omit(cows)
missingsTable <- 10 - table(cows.com$dose, cows.com$time)
print(xtable(missingsTable, digits = 0, caption = "Number of missing values in PCV for time and dose"), comment = FALSE, , table.placement = "H")
```

It seems to be a contradiction with the assumption that missing data are randomly distributed. In the study all those missing values will be removed and will not be further analyzed nor imputed.

## Correlation structure

```{r correlation, results = 'asis'}
library("tidyr")
cows.w <- spread(cows, time, pcv)
print(xtable(cor(cows.w[, c("1", "2", "3")], use = "pairwise.complete.obs"), caption = "Time correlation. \\label{tab:correlation}"), comment = FALSE, table.placement = "H", include.rownames = FALSE)
```

This correlation structure was obtained for the time. A positive correlation can be observed in Table \ref{tab:correlation}. This means that a possible correlation structure can be applied to the data. One could use an autoregressive model AR(1) or a Compound Symmetry that implies that all distinct members of a cluster are equally correlated with each other. The total variation can be separated between the variation within cluster and the one between cluster. 


# Results

Each cow in this dataset was treated three times, hence we will, at the beginning at least, consider that our dataset contains 30 different cows identified by id and dose. Later on this assumption will be released. The idea behind this differentiation within the same cow is due to the fact that a different Berenil dose was given for each infection. That is to say each cow has been infected three different times by the parasite.

## Multivariate model


```{r multivariate_model, echo=FALSE, results='hide'}
round.pval <- function(pval=stop("p-value not given"), show.sig.stars=FALSE, alpha=0.05){
  
  arrodonir <- function(pval, show.sig.stars, alpha){
    
		if(is.na(pval)) return(NA)

		ndec <- nchar(as.character(signif(alpha,1)))-2
		pval2 <- if(pval >= alpha) round(pval, ndec) else signif(pval, 1)

		i <- 1
		while(pval2==signif(alpha, i)){
			pval2 <- round(pval, ndec+i)
			if(pval2==pval) break
			i <- i+1
		}
		pval2 <- format(pval2, scientific=F)
		if(pval<0.00001){
			pval2 <- "<0.00001"
		}
		
		if(show.sig.stars){
			cpoints <- c(0, alpha/50, alpha/5, alpha, ceil(alpha*10^(ndec-1))/10^(ndec)+alpha, 1)
			c(0, 0.001, 0.01, 0.05, 0.1, 1)
			pval2 <- paste(pval2, symnum(pval, corr = FALSE, na = FALSE, cutpoints = cpoints, symbols = c("***", "**", "*", ".", " ")))
		}

		return(pval2)
	}

	ret <- sapply(pval, function(x) arrodonir(x, show.sig.stars=show.sig.stars, alpha=alpha))
	if(is.matrix(pval)) ret <- matrix(ret, ncol=ncol(pval))
	return(ret)

}

cows <- read.table("data/cattle_mes dades.txt", header = TRUE, 
                   sep = "\t", dec = ",", na.strings = "")

names(cows) <- c("id", "dose", "pcv", "time", "nbirth")

cows$id <- as.factor(cows$id)
cows$dose <- factor(cows$dose, levels=c("L", "M", "H"))

# dep variable:
# pcv: the higher the better.

# indep variables:
# dose: we want to know if the dose is associated with lower PCV
# nbirth: we want to know if this covariate influences the effect of the dose.
# time: we dont particularly want to see differences in time, but maybe 
# 3rd observations of the doses are higher, ...

# we do not want:
# id: we dont want the effect of the dose to be explained by the cow.

# Each observation of each cow will be considered independent.
taula <- matrix(ncol=6, nrow=19)
model <- lm(pcv ~ time + dose, data=cows)
summary(model)
model0 <- model
colnames(taula) <- c("Formula", "Variables", colnames(summary(model)$coef))
as.character(model$terms)->formu
taula[1,1] <- paste(formu[c(2,1,3)], collapse="")
taula[1:4,-(1:2)] <-  summary(model)$coef
taula[1:4,2] <- rownames(summary(model)$coef)

model <- lm(pcv ~ time*dose, data=cows)
summary(model)
model1 <- model
as.character(model$terms)->formu
taula[6,1] <- paste(formu[c(2,1,3)], collapse="")
taula[6:11,-(1:2)] <-  summary(model)$coef
taula[6:11,2] <- rownames(summary(model)$coef)
# dose H has a huge effect, in a possitive way. Also, in general higher
# times give better PCV values (which is logical). nbirth has a significative
# negative effect on the PCV.

model <- lm(pcv ~ time*dose + nbirth, data=cows)
summary(model)
model2 <- model
as.character(model$terms)->formu
taula[13,1] <- paste(formu[c(2,1,3)], collapse="")
taula[13:19,-(1:2)] <-  summary(model)$coef
taula[13:19,2] <- rownames(summary(model)$coef)
# time and dose almost interact for high doses. This interaction is synergistic.
# higher times and dose H give even higher pcv, but this effect is higher when 
# combined together.

model <- lm(pcv ~ dose + time*nbirth, data=cows)
summary(model)
as.character(model$terms)->formu
model <- lm(pcv ~ time + dose*nbirth, data=cows)
summary(model)
# no other interactions are significative.
taula[,6] <- round.pval(as.numeric(taula[,6]))
taula[,3:5] <- round(as.numeric(taula[,3:5]), 2) 
# print(xtable(taula, digits=c(0,0,0,2,2,2,2)), include.rownames=F)
# so... model selection, for the more pasimonious model says...
add1(model0, scope=~.^2+nbirth, test="F")
add1(update(model0,~.^2), scope=~.+nbirth, test="F")
add1(model2, scope=~.^2, test="F")

## final model:
model2 <- lm(pcv ~ time*dose + nbirth, data=cows, y=T, model=T)
summary(model2)
anova(model2)
```

Table \ref{taulamodels} summarizes the results of the estimation of the parameters of all models. The model selection which was used is the forward step-wise method. The last model corresponds to the most parsimonious one. It is the one that better explains the data, here is its specification:

$$
PCV= 15.82 + 2.43X_{time} -0.32X_{doseM} -1.31X_{doseH} -0.34X_{nbirth} + 0.48X_{doseM}X_{time} + 2.35X_{doseH}X_{time}
$$

Although the coefficients for the additive effects for both doses are negative, when adding the interaction term it ensures a positive effect. The difference in PCV for the same cows treated with high dose with respect to low dose will be:

$$
PCV(X_{doseH}=1)-PCV(X_{doseL}=1) = -1.31+2.35X_{time} 
$$

Notice that for every time the result is positive. Therefore we can consider that the high dose has a bigger effect than low dose (both coefficients are significantly different from 0). For the medium dose the coefficients are not significant. These calculations are always with respect to the same cow, when being treated with different doses.

$$
PCV(X_{doseM}=1)-PCV(X_{doseL}=1) = -0.32+0.48X_{time} 
$$

Here the effect of the dose changes with the time in a synergistic way. 
Also, time has a positive effect in the PCV(p$<$0.00001). However the number of births seems to have a significant negative effect in the PCV (p=0.01).



\begin{table}[H]
\centering
\begin{tabular}{llllll}

\hline
Formula & Variables & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
pcv $\sim$ time + dose & (Intercept) & 12.32 & 0.83 & 14.78 & $<$0.00001 \\ 
   & time & 3.55 & 0.37 & 9.63 & $<$0.00001 \\ 
   & doseM & 0.56 & 0.75 & 0.75 & 0.45 \\ 
   & doseH & 3.2 & 0.72 & 4.47 & 0.00003 \\ 
   &  &  &  &  &  \\ 
  pcv $\sim$ time * dose & (Intercept) & 14.51 & 1.29 & 11.28 & $<$0.00001 \\ 
   & time & 2.28 & 0.68 & 3.34 & 0.001 \\ 
   & doseM & -0.47 & 1.81 & -0.26 & 0.79 \\ 
   & doseH & -1.45 & 1.73 & -0.84 & 0.41 \\ 
   & time:doseM & 0.65 & 0.93 & 0.7 & 0.48 \\ 
   & time:doseH & 2.5 & 0.87 & 2.87 & 0.005 \\ 
   &  &  &  &  &  \\ 
  pcv $\sim$ time * dose + nbirth & (Intercept) & 15.85 & 1.35 & 11.74 & $<$0.00001 \\ 
   & time & 2.43 & 0.66 & 3.67 & 0.0005 \\ 
   & doseM & -0.32 & 1.75 & -0.18 & 0.86 \\ 
   & doseH & -1.31 & 1.67 & -0.78 & 0.44 \\ 
   & nbirth & -0.34 & 0.13 & -2.51 & 0.01 \\ 
   & time:doseM & 0.48 & 0.9 & 0.54 & 0.59 \\ 
   & time:doseH & 2.35 & 0.84 & 2.8 & 0.007 \\ 
   \hline
\end{tabular}
\caption{All multivariant models considered for fitting the data that were significantly different, using F-tests to assess this difference.}
\label{taulamodels}

\end{table}

The analyses of the residuals of the model can be seen in Figure \ref{resplots}. Clearly, they present some issues: residuals are not independent from PCV, they are not centered for the high dose, there is no homoscedasticity, etc. Even though this model is not optimal model to fit the data (and this is also shown in the residuals plots), it is pointing in a pretty obvious way that the high dose is the only one that an effect significantly different from low dose. One of the problems with the data is that there is no control among the cows so the effect of the doses cannot be compared with PCV without any treatment.


```{r validation_multi, fig.cap='Analyses of the residuals of the model.\\label{resplots}', fig.pos = "H"}
v1 <- qplot(model2$y, model2$residuals, xlab = "PCV value", ylab = "Residuals")
v1 <- v1 + geom_hline(yintercept=0) + geom_smooth()
v2 <- qplot(model2$model$dose, model2$residuals, ylab="Residuals", xlab = "Dose", geom = "boxplot", fill = model2$model$dose) + guides(fill = FALSE) + scale_x_discrete(labels = c("Low", "Medium", "High"))
v3 <- qplot(model2$model$time, model2$residuals, ylab = "Residuals", xlab = "Time") + 
  scale_x_discrete(labels = 1:3) + geom_hline(yintercept=0)
v4 <- qplot(model2$model$nbirth, model2$residuals, ylab = "Residuals", xlab = "Number of births (nbirth)") +
  geom_hline(yintercept=0)
grid.arrange(v1, v2, v3, v4, ncol = 2)
```



## Two stage analysis 

In this section a two stage analysis is performed. A linear regression is fitted for each cow in the first stage. For this purpose a new variable _idDose_ was created merging id and dose, so that 30 different cows could be considered. Then, with the coefficients obtained, a second linear regression is done for the intercept and the slope conditioning on the dose.

Several issues emerge with the missing values. As it was specified in the missing data analysis section, several cows have just a pcv value for time 1, so it was not possible to generate a slope. For instance cows number 6 and 9 treated with low dose do not have a slope after stage one.. 

__Stage 1 Model__

$PCV_{ij} = \beta_{0ij} + \beta_{1ij}time + \epsilon_{ij}$ and $\epsilon_{ij} \sim N(0, \Sigma_ij)$


__Stage 2 Model__

$\beta_{0ij} = \beta_0 + \beta_1 doseMedium + \beta_2 doseHigh + b_{0ij}$

$\beta_{1ij} = \beta_3 + \beta_4 doseMedium + \beta_5 doseHigh + b_{1ij}$

where $b_i \sim N(0, D)$

```{r}
cows.com$idDose <- paste(cows.com$id, cows.com$dose, sep = "_")
cows.gd <- groupedData(pcv ~ time|idDose, data = cows.com, 
                       outer = ~dose, inner = ~nbirth)

cows.lmList <- nlme:::lmList(pcv ~ time, cows.gd)
betas <- as.data.frame(coef(cows.lmList))
names(betas) <- c("Intercept", "slope")

betas <- add_rownames(betas, var = "idDose")
betas.info <- cows.com %>% group_by(idDose, id, dose) %>% summarize(nbirth = mean(nbirth))
bdd <- left_join(betas, betas.info)

# modbeta0 <- lm(Intercept ~ dose, bdd)
# modbeta1 <- lm(slope ~ dose, bdd)
# summary(modbeta0)
# summary(modbeta1)
```

Figure \ref{fig:two_stage_graph} shows all the linear regressions done at the first stage analysis. The slopes are positive for each dose but bigger for the high dose treatment. By comparing this Figure with Figure \ref{fig:qplot} one can notice that a lot of information about the variability is lost.

```{r two_stage_graph, fig.cap = "Each line represents the linear model fited for each cow and dose.\\label{fig:two_stage_graph}", fig.pos = "H"}
# Graph option 1
# q <- qplot(factor(time), pcv, group = id, data = cows.com, geom = c("point", "line")) + facet_grid(dose ~ id)
# q + geom_abline(aes(intercept = Intercept, slope = slope), colour = "red", data = bdd, alpha = 0.8)

bdd$doseExt <- factor(bdd$dose, labels = c("Low", "Medium", "High"))
 
# Graph option 2
q <- qplot(factor(time), pcv, group = id, data = cows.com, geom = "point") + facet_grid(.~ doseExt)
q + geom_abline(aes(intercept = Intercept, slope = slope, colour = factor(doseExt)), data = bdd) + xlab("time") + ylab("PCV value")
```


\begin{table}[H]
\centering
\begin{tabular}{lll}

\hline
Parameters            & Estimate (p-value)       \\ 
\hline
$\beta_{0}$ (Intercept)  & 15.1300 (\textless 2e-16) \\
$\beta_{1}$ (doseMedium) & -0.6448 (0.5237)         \\
$\beta_{2}$ (doseHigh)   &  -2.0667 (0.0429)   \\   
\hline
\end{tabular}
\caption{Two-stage analysis results of model fitting of $\beta_{0ij}$}
\label{tab::Two_stage_intercept}
\end{table}

The parameter estimates for $\beta_{0ij}$ are included in Table \ref{tab::Two_stage_intercept} along with their corresponding p-value. The intercept (15.1300) represents the mean PCV value at time 1 for a cow treated with low dose, as it is the reference treatment. The parameter estimate for the medium dose is not significant. However the estimate for dose high (-2.0667) is significant and it means that the intercept for a cow treated with dose high would be 13.0633.


\begin{table}[H]
\centering
\begin{tabular}{ll}
\hline
Parameters   & Estimate (p-value)       \\ 
\hline
$\beta_{3}$ (Intercept)   & 2.1500 (0.000438)  \\
$\beta_{4}$ (doseMedium) &  0.4500 (0.540909)  \\
$\beta_{5}$ (doseHigh)  &  2.6300 (0.001083)  \\
\hline
\end{tabular}
\caption{Two-stage analysis results of model fitting of $\beta_{1ij}$}
\label{tab::Two_stage_slope}
\end{table}

Here we have the estimates and p-values for $\beta_{1ij}$. The intercept (2.1500) represents the average slope. As it was seen in the previous table, the medium dose is not significant with respect to the low dose. However, the high dose is strongly significant with an estimate of 2.6300 which means that the slope for a cow treated with high dose would be 4.78.

## Random effects

Since variability is lost in the Two stage analysis, a linear mixed model is performed. It allowed us to add the variability due the difference between cows identified by the variable _idDose_ which merges _id_ and _dose_.

### Mixed effect model with _idDose_

As explained in the section \ref{Statistical_methods:RE}, a starting model was fitted. Then several alternative models including a random effect were fitted. Table \ref{anovas:Random_slope_intercept} summarizes the model selection process for the selection of a random effects structure.

```{r, results='hide'}
# Fixed effect
cows.gls <- gls(pcv ~ time + dose, data = cows.com, method = "REML")

# Random effect
cows.lme.Inte <- lme(pcv ~ time + dose, random = ~1|idDose, data = cows.com)
cows.lme.Slop <- lme(pcv ~ time + dose, random = ~0 + time|idDose, data = cows.com)

taulainte <- as.matrix(anova(cows.gls, cows.lme.Inte))[,-1]
taulainte <- taulainte[,-c(1, 6)]

taulaslope <- as.matrix(anova(cows.gls, cows.lme.Slop))[,-1]
taulaslope <- taulaslope[,-c(1, 6)]

taula <- taulainte
taula <- rbind(taula, taulaslope[2,])
taula <- apply(taula,1:2,as.numeric)
taula[,-6] <- round(taula[,-6],3)
rownames(taula) <- c("Fixed Effects Model", "Random Intercept Model", "Random Slope Model")
taula[-1,6] <- round.pval(taula[-1,6])

print(xtable(taula[,-(3:4)], label="anovas:Random_slope_intercept", caption="Anova table comparing the Fixed effects model (no random effects) with the random intercept and the random slope models."), comment = FALSE, table.placement = "H")
```

\begin{table}[H]
\centering
\begin{tabular}{rrrrr}
  \hline
 & df & AIC & L.Ratio & p-value \\ 
  \hline
$\beta_{0} + \beta_{1}time + \beta_2 doseMedium + \beta_3 doseHigh$ & 5 & 361.32 &  &  \\ 
 $(\beta_{0} + b_{0i}) +  \beta_{1}time + \beta_2 doseMedium + \beta_3 doseHigh$ & 6 & 352.25 & 11.07 & 0.0009 \\ 
 $\beta_{0} +  (\beta_{1} + b_{1i})time + \beta_2 doseMedium + \beta_3 doseHigh$ & 6 & 329.37 & 33.95 & <0.00001 \\  
   \hline
\end{tabular}
\caption{Anova table comparing the Fixed effects model (no random effects) with the random intercept and the random slope models.} 
\label{anovas:Random_slope_intercept}
\end{table}

Table \ref{anovas:Random_slope_intercept} includes different models with their comparison against the reference model. An information criterion (AIC) and the corresponding p-value is also included. Model 0 was the starting and reference model with only fixed effects. At model 1 a random effect was added at the time and for model 2 it was added at the intercept.

The models that included the random effect were significantly better than the reference one. In order to compare the two remaining models, the AIC was used because they were not nested and a LRT could not be performed. In this case model 2 was preferred as having lower AIC.

In following steps variable nbirth was included in the model (p = 0.04). Neither adding of random effects nor including a correlation structure was significant. In order to have the most parsimonious model the following one was selected:

$$PCV_{i} = \beta_{0} + (\beta_{1} + b_{1i})time + \beta_2 doseMedium + \beta_3 doseHigh + \beta_4 numbirths + \epsilon_{i}$$

```{r table_lme_parametrs, results='asis'}
cows.lme.Slop <- update(cows.lme.Slop, method = "ML")
cows.lme.Slop.birth <- update(cows.lme.Slop, fixed  = ~ . + nbirth, method = "ML")
cows.lme.Slop.birth <- update(cows.lme.Slop.birth, method = "REML")
intr <- intervals(cows.lme.Slop.birth)

val.brackets <- function(valuesvect, lower=NULL, upper=NULL, between="-"){
  
  return(
    if(!is.null(upper) & !is.null(lower)) paste(valuesvect, " (", lower, between, upper, ")", sep="")
    else if(is.null(upper) & !is.null(lower)) paste(valuesvect, "(", lower, ")", sep="")
    else if(is.null(upper) & is.null(lower)) paste(valuesvect[,1], " (", valuesvect[,2], between, valuesvect[,3], ")", sep="")
  )
}

taula.intervals <- matrix(nrow=10, ncol=3)
colnames(taula.intervals) <- c(" ", "Parameter", "Estimate (95% CI)")

taula.intervals[1,1] <- "Fixed Effects:"
taula.intervals[2:6,2] <- rownames(intr$fixed)
taula.intervals[2:6,3] <- val.brackets(round(intr$fixed[,c(2,1,3)],2), between=", ")
taula.intervals[7,1] <- "Random Effects:"
taula.intervals[8,2] <- "idDose:"
taula.intervals[9,2] <- rownames(intr[[2]]$idDose)
taula.intervals[9,3] <- val.brackets(round(intr[[2]]$idDose[c(2,1,3)],2), between=", ")
taula.intervals[10,1] <- "Within-group standard error:"
taula.intervals[10,3] <- val.brackets(round(intr[[3]][2],2), round(intr[[3]][1],2), round(intr[[3]][3],2), between=", ")


sumslope <- summary(cows.lme.Slop.birth)
taula.sumslope <- matrix(nrow=10, ncol=7)
colnames(taula.sumslope) <- c("", "Parameters", colnames(sumslope$tT))
taula.sumslope[2:6,2] <- rownames(sumslope$tT)
taula.sumslope[1,1] <- "Fixed Effects:"
taula.sumslope[2:6,3:6] <- round(sumslope$tT[,-5],3)
taula.sumslope[2:6,7] <- round.pval(sumslope$tT[,5])

taula <- cbind(taula.intervals, taula.sumslope[, 4:7])
print(xtable(taula, label="intervals", caption="Summary of the estimations for the parameters for the random slope model adding the number of births. \\label{tab:table_lme_parametrs}"), include.rownames = FALSE, comment = FALSE, table.placement = "H")

```

The parameter estimates for the model are included in Table \ref{tab:table_lme_parametrs} along with their corresponding p-value. The intercept (14.69) represents the mean PCV value at time 1 for a cow treated with low dose. The parameter estimates for the medium dose and high dose are not significant. This result disagree with all previous data exploration and models. It seems that using the model simplification with _idDose_ does not take into account _id_ variability.

Figure \ref{slope_graph} shows the mean slope $\beta_{1} = 3.55$. The random effect $b_{1i} = 1.06$ can be seen in the grey part of the graph. For instance _idDose_ slopes are present for cows 5_L, 8_H, 9_M.


```{r slope_graph, fig.cap = "Fixed and random effect of time, with some examples. \\label{slope_graph}", fig.pos = "H", fig.height = 4, fig.width = 6}
cows.lme.Slop.birth <- lme(pcv ~ time + dose + nbirth, random = ~0 + time|idDose, data = cows.com)
cows.fixef <- fixed.effects(cows.lme.Slop.birth)
cows.ranef <- ranef(cows.lme.Slop.birth, augFrame = TRUE)[c(1, 17, 29, 24),]
cows.ranef$intercept <- cows.fixef[1]
cows.ranef <- add_rownames(cows.ranef, "idDose")
cows.ranef$yend <- cows.ranef$intercept + 3*(cows.ranef$time + cows.fixef[2])
cows.ranef$yend[1] <- cows.fixef[1] + 3*cows.fixef[2]
cows.ranef$idDose[1] <- "Fixed Effect"
bbd <- data.frame(x = c(1, 3, 3), y = cows.fixef[1] + c(0, 3*(cows.fixef[2] + 2*1.062889), 3* (cows.fixef[2] - 2*1.062889)))

q <- ggplot(aes(factor(time), pcv), data = cows.com) + geom_blank() + 
  ylim(c(10, 33))
q <- q + geom_polygon(aes(x = x, y = y), alpha = 0.2, data = bbd) + xlab("time") + ylab("PCV value") 
q <- q + geom_segment(aes(x = 1, xend = 3, y = intercept, yend = yend, colour = idDose), size = 2, cows.ranef)
q + xlab("time") + ylab("PCV value") + theme(legend.title=element_blank())
```

As shown in Figure \ref{slope_graph} $b_{1i}$ is the variability added to the mean slope in time depending on each _idDose_. It means that the effect of time is different for each _idDose_. For instance, if a given cow had a positive random effect, time would have a bigger slope. Whatever the value $b_{1i}$ is, time has always a positive impact on PCV.

### Mixed effect model with id and nested dose

```{r lme_nested}
cows.lme.nested <- lme(pcv ~ time + dose + nbirth, random = ~0 + time|id/dose, data = cows.com)
```

The model selected in the previous section was kept. Whereas the grouping was initially performed on _idDose_ it is now done on _id_ then _dose_. The point is to get the variability for each id then the variability within each _id_ for each dose. In this way we are getting a random effect for _id_ and for _dose_ in _id_. The second random effect differences the effect of time within each cow depending on the kind of doses that was administrated.

$PCV_{ij} = \beta_{0} + (\beta_{1} + b_{1i} + b_{1ij})time + \beta_2 doseMedium + \beta_3 doseHigh + \beta_4 numbirths + \epsilon_{i}$ and $\epsilon_{i} \sim N(0, \Sigma_ij)$ i stands for id, j stands fore dose.

The parameters estimates of this model do not change much from previous model. Basically the estimate for medium dose is 0.297 (p = 0.6680) and 1.671 (p = 0.021) for high dose. With this model the high dose is significant and has a positive effect on pcv.

Figure \ref{random_graph} shows the random effect of each dose within the cows.

```{r random_graph, fig.cap= "Random effect for time of dose in each id. \\label{random_graph}", fig.height = 4, fig.width = 6, fig.pos = "H"}
randomEffects <- ranef(cows.lme.nested)
ranEfDf <- randomEffects$dose
ranEfDf <- add_rownames(ranEfDf, "idDose")
ranEfDf <- separate(ranEfDf, idDose, c("id", "dose"), remove = FALSE)
ranEfDf$dose <- factor(ranEfDf$dose, levels = c("L", "M", "H"))
ranEfDf$doseExt <- factor(ranEfDf$dose, labels = c("Low", "Medium", "High"))
qplot(time, id, data = ranEfDf, color = doseExt, size = I(4)) + theme(legend.title=element_blank()) +
  xlab("Random effect for time")
```

Those random effects modify the effect of time for each cow. The slope of time is modified by the random effect for the _id_ and then by the random effect for the _dose_. Figure \ref{random_graph} shows these values. A positive value means that, within the cow, the considered dose has a positive effect on the mean slope time, that is to say on the effect of time on PCV. One can notice that the high dose almost always has a value superior to the medium or low dose. Once again the efficiency of the high dose is shown to be superior to the one of the other doses. 

Whenever the value of the random effect for the high dose is lower than the one for the other doses it can be explain by the data. In fact it is either due to missing values or a really specific cow. For instance cow 4 has a lower PCV value after time 2 for the high dose. The validation of this model can be seen in the appendix (Figure \ref{nesteddiagnostic}).

# Discussion

TODO(All): Revise comments

The main objective of this study was to assess the efficiency of the different kinds of doses used to cure cows infected with the trypanosomosis parasite. As expected _time_ and _dose_ have a positive effect on PCV. However the effect of the dose varies with the kind of doses used. When comparing doses to the low dose as a reference it appears that the medium dose has not an effect significantly different from the low one. However the high dose is much more efficient. This is true when analysing the mean effect in a multivariate model and also for each cow when performing a linear mixed model.

Giving birth is supposed to weaken the cow, besides *nbirth* is also an indicator of how old the cow is. Hence we should find a negative relation between the number of births and the PCV. First such a relation did not appear clearly in the descriptive analysis but models showed that indeed the number of births has a negative impact on the PCV.

# Limitations and Further research

The relative low number of observations in the dataset makes difficult the analysis because as soon as a cow has particular results it influences a lot the results. A lot of assumptions were needed to study the data. In order to verify whether or not those assumptions can be validated we would have needed more information on the dataset. For instance the reason why values are missing could have helped us give a better interpretation of the results of the random effects of the linear mixed model. 

To clarify the dataset and in a way assumptions are no longer needed, one can contact the person who collected the data. Adding some covariates may also help futur researchers predict PCV values based on the initial condition of a given cow.
In order to further analyse the data some improvements are possible. Missing values can be imputed so that information is no longer lost in some part of the study. One can also try to find a proper correlation structure to add in the differents models. 

TODO(Mathieu):
We did not explore interaction between time and dose in the linear mixed models.

* May be there is some selection bias. Cows from other regions are not properly modeled. Need more data.
* We de not have a control. We cannot estimate if some dosage is not needed.
* It is not clear whici PCV value needed for a cow to be cured.


TODO(All): Revise comments

# Bibliography

* [Bovine trypanosomiasis in south-western Uganda: packed-cell volumes and prevalences of infection in the cattle.](http://www.ncbi.nlm.nih.gov/pubmed/15000727)
* [Evaluating the use of packed cell volume as an indicator of trypanosomal infections in cattle in eastern Zambia](http://www.sciencedirect.com/science/article/pii/S0167587708001098#)
* [Linear Mixed Models for Longitudinal Data](http://www.springer.com/gp/book/9781441902993)


# Appendix

```{r validacio_nested, echo=FALSE, fig.cap='Analyses of the residuals of the nested model.\\label{nesteddiagnostic}', fig.pos = "H"}
model2 <- cows.lme.nested
model2$residuals <- predict(model2)-cows.com$pcv
v1 <- qplot(cows.com$pcv, model2$residuals, xlab = "PCV value", ylab = "Residuals")
v1 <- v1 + geom_hline(yintercept=0) + geom_smooth(span = 1.2)
v2 <- qplot(model2$data$dose, model2$residuals, ylab="Residuals", xlab = "Dose", geom = "boxplot", fill = model2$data$dose) + guides(fill = FALSE) + scale_x_discrete(labels = c("Low", "Medium", "High"))
v3 <- qplot(model2$data$time, model2$residuals, ylab="Residuals", xlab="Time") + 
  scale_x_discrete(labels = 1:3) + geom_hline(yintercept=0)
v4 <- qplot(model2$data$nbirth, model2$residuals, ylab = "Residuals", xlab = "Number of births (nbirth)") + geom_hline(yintercept=0)
grid.arrange(v1, v2, v3, v4, ncol = 2)
```

TODO: details more or less important
* first page should only contain the title and our names.
* add some space between section.
* figure 2 is misplaced, it should be before the paragraph before number of births.
* table 3 should be after the model (the one after the first paragraph of multivariate model)
* figure 3 should be before the paragraph before two stage analysis.
* table 4 and 5 should ?aybe be after the paragraphs that give the explaination since we do not introduce the tables.
* figure 4 should be before the paragraph before Mixed effect model with id and nested dose.
* figure 5 should not be in middle of the paragraph.
* Discussion and Limitation and further research should become one section COnclusion. It really weird to not have a conclusion section.